Concurrent Programming with CXL

### 2020-06-29 book/paper.tex
Made a formatting pass over all chapters

As a first example, we will revisit the producer/consumer problem.
The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \texttt{items} is an unbounded array with each
entry initialized to \texttt{()}.  \texttt{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc(p)}: atomically increments \texttt{\^{}p} and returns
the old value;
\item \texttt{swap(p)}: sets \texttt{\^{}p} to \texttt{()} and returns
the old value.
\end{itemize}

Method \texttt{produce(item)} uses \texttt{inc(\&back)} to allocate
the next available slot in the \texttt{items} array.
It stores the item as a singleton tuple.
Method \texttt{consume()} repeatedly scans the array, up to the
\texttt{back} index, trying to find an item to return.
To check an entry, it uses \texttt{swap()}
to atomically remove an item from a slot if there is one.
This way, if two or more processes attempt to extract an item from
the same slot, at most one will succeed.

There is no critical section.  If one process is executing instructions
very slowly, this does not negatively impact the other processes, as it
would with solutions based on critical sections.
On the contrary, it helps them because it creates less contention.
Unfortunately, the solution is not practical for the following reasons:
\begin{itemize}
\item The \texttt{items} array must be of infinite size if an unbounded number
of items may be produced;
\item Each slot in the array is only used once, which is inefficient;
\item the \texttt{inc} and \texttt{swap} interlock instructions are not
universally available on existing processors.
\end{itemize}
However, in the literature there are many examples of practical
non-blocking (aka \emph{wait-free})
\index{wait-free synchronization}%
synchronization algorithms~\cite{}.

@@

The list has two ``book-end'' nodes with values \texttt{-inf} and
\texttt{inf} (like the Python \texttt{math.inf} constants).
Each node has a lock, a value, and \texttt{next}, a pointer to the next node
(which is \texttt{None} for the final \texttt{inf} node).
The \texttt{lst\_find(lst, v)} method is a helper function for update
methods that
finds and locks two consecutive nodes \texttt{before} and \texttt{after}
such that \texttt{before.value $<$ v $\le$ after.value}.
It does so by performing something called \emph{hand-over-hand locking}.
\index{hand-over-hand locking}%
It first locks the first node, which is the \texttt{-inf} node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a tree
hand-over-hand.

Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
and \texttt{lst\_remove} is straightforward.
The \texttt{lst\_contains} method can simply scan through the list
because the \texttt{next} pointers are updated atomically in such a
way that they always point to a legal suffix of the list.
An invariant of the algorithm is that at any point in time the
list is ``complete'', starting with a \texttt{-inf} node and ending
with a \texttt{inf} node.

Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization points for the \texttt{lst\_insert} and
\texttt{lst\_remove} operations coincide exactly with the update
of the \texttt{before.next} pointer.
The linearization point of a \texttt{lst\_contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
\texttt{inf} node.

@@

\section*{Exercises}
\begin{problems}
\item Add read-only methods to the data structure in Figures~\ref{fig:lst1}
and~\ref{fig:lst2} that report the size of the list, the minimum value in the
list, the maximum value in the list, and the sum of the values in the list.
Are they linearizable?  If so, what are their linearization points?
\item A \emph{seqlock}
\index{seqlock}%
consists of a lock and a version number.
An update operation acquires the lock, increments the version number, makes the
changes to the data structure, and then releases the lock.  A read-only operation
does not use the lock.  Instead, it retrieves the version number,
reads the data structure, and then checks if the
version number has not changed.  If it has, read-only operation is retried.
Use a seqlock to implement a bank much like Exercise~\ref{ex:bank}, with
one seqlock for the entire bank (i.e., no locks on individual accounts).
Method \texttt{transfer} is an update operation; method \texttt{total} is a
read-only operation.  Explain how a \texttt{seqlock} can lead to starvation.
\end{problems}

@@

The \texttt{round} array is kept to check the correctness of this
approach.  Each process increments its entry every time it enters
a round.  If the algorithm is correct, it can never be that two processes
are more than one round apart.

\section*{Exercises}
\begin{problems}
\item See if you can implement barrier synchronization for $N$ processes
with just three semaphores.  Busy waiting is not allowed.  How about just two?
(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
for solving semaphore problems.  Look for the double turnstile solution.)
\item Implement barrier synchronization with Mesa condition variables.
(You may want to use a double turnstile approach here as well.)
\end{problems}

>>>>

As a first example, we will revisit the producer/consumer problem.
The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \textit{items} is an unbounded array with each
entry initialized to \texttt{()}.  \textit{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc}($p$): atomically increments \texttt{\^{}}$p$ and returns
the old value;
\item \texttt{swap}($p$): sets \texttt{\^{}}$p$ to \texttt{()} and returns
the old value.
\end{itemize}

Method \texttt{produce}(\textit{item}) uses \texttt{inc}(\&\textit{back})
to allocate
the next available slot in the \textit{items} array.
It stores the item as a singleton tuple.
Method \texttt{consume()} repeatedly scans the array, up to the
\textit{back} index, trying to find an item to return.
To check an entry, it uses \texttt{swap()}
to atomically remove an item from a slot if there is one.
This way, if two or more processes attempt to extract an item from
the same slot, at most one will succeed.

There is no critical section.  If one process is executing instructions
very slowly, this does not negatively impact the other processes, as it
would with solutions based on critical sections.
On the contrary, it helps them because it creates less contention.
Unfortunately, the solution is not practical for the following reasons:
\begin{itemize}
\item The \textit{items} array must be of infinite size if an unbounded number
of items may be produced;
\item Each slot in the array is only used once, which is inefficient;
\item the \texttt{inc} and \texttt{swap} interlock instructions are not
universally available on existing processors.
\end{itemize}
However, in the literature there are many examples of practical
non-blocking (aka \emph{wait-free})
\index{wait-free synchronization}%
synchronization algorithms~\cite{}.

@@

The list has two ``book-end'' nodes with values \texttt{-inf} and
\texttt{inf} (like the Python \texttt{math.inf} constants).
Each node has a lock, a value, and \textit{next}, a pointer to the next node
(which is \texttt{None} for the final \texttt{inf} node).
The \texttt{lst\_find}(\textit{lst}, $v$) method is a helper function for update
methods that
finds and locks two consecutive nodes \textit{before} and \textit{after}
such that $\mathit{before}.\mathtt{data.value} < v \le \mathit{after}.\mathtt{data.value}$.
It does so by performing something called \emph{hand-over-hand locking}.
\index{hand-over-hand locking}%
It first locks the first node, which is the \texttt{-inf} node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a tree
hand-over-hand.

Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
and \texttt{lst\_remove} is straightforward.
The \texttt{lst\_contains} method can simply scan through the list
because the \textit{next} pointers are updated atomically in such a
way that they always point to a legal suffix of the list.
An invariant of the algorithm is that at any point in time the
list is ``complete'', starting with a \texttt{-inf} node and ending
with a \texttt{inf} node.

Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization points for the \texttt{lst\_insert} and
\texttt{lst\_remove} operations coincide exactly with the update
of the \textit{before}.\texttt{next} pointer.
The linearization point of a \texttt{lst\_contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
\texttt{inf} node.

@@

\section*{Exercises}
\begin{problems}
\item Add read-only methods to the data structure in Figures~\ref{fig:lst1}
and~\ref{fig:lst2} that report the size of the list, the minimum value in the
list, the maximum value in the list, and the sum of the values in the list.
Are they linearizable?  If so, what are their linearization points?
\item A \emph{seqlock}
\index{seqlock}%
consists of a lock and a version number.
An update operation acquires the lock, increments the version number, makes the
changes to the data structure, and then releases the lock.  A read-only operation
does not use the lock.  Instead, it retrieves the version number,
reads the data structure, and then checks if the
version number has not changed.  If it has, read-only operation is retried.
Use a seqlock to implement a bank much like Exercise~\ref{ex:bank}, with
one seqlock for the entire bank (i.e., no locks on individual accounts).
Method \texttt{transfer} is an update operation; method \texttt{total} is a
read-only operation.  Explain how a seqlock can lead to starvation.
\end{problems}

@@

The \textit{round} array is kept to check the correctness of this
approach.  Each process increments its entry every time it enters
a round.  If the algorithm is correct, it can never be that two processes
are more than one round apart.

\section*{Exercises}
\begin{problems}
\item See if you can implement barrier synchronization for \texttt{N} processes
with just three semaphores.  Busy waiting is not allowed.  How about just two?
(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
for solving semaphore problems.  Look for the double turnstile solution.)
\item Implement barrier synchronization with Mesa condition variables.
(You may want to use a double turnstile approach here as well.)
\end{problems}
